---
output: github_document
---

<!-- Render with make head-to-head; the drivers write the JSON receipts. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
up <- jsonlite::read_json("head-to-head-upstream.json")
ours <- jsonlite::read_json("head-to-head-ours.json")
threads <- jsonlite::read_json("threading.json")
before <- jsonlite::fromJSON(paste(system2("git", c("show",
  "71c4674:benchmarks/head-to-head-ours.json"), stdout = TRUE), collapse = "\n"), simplifyVector = FALSE)
timed <- function(runs) Filter(function(r) !isTRUE(r$warmup), runs)
stat <- function(x) sprintf("%.3f [%.3f–%.3f]", median(x), min(x), max(x))
col <- function(runs, f) vapply(timed(runs), f, numeric(1))
measured <- unique(vapply(ours, `[[`, "", "commit"))
stopifnot(length(measured) == 1L, !any(vapply(ours, `[[`, FALSE, "dirty")))
```

# Head-to-head: duckpeakwhere vs upstream peakwhere

W2 on the same 13th Gen Intel Core i5-13500 (20 logical CPUs), headless Chromium
148.0.7778.96. One warm-up plus five measured app runs, each in a fresh browser.
Timing jobs ran sequentially, without another test or benchmark job launched alongside
them. Resident system and agent services remained; this was not an exclusive host.

- **Upstream:** [seandavi/peakwhere](https://github.com/seandavi/peakwhere) at
  `f40a673842815b9eef08c4d32243b29dac3e0646`, using its committed application and real
  file inputs. The upstream receipt is the committed 2026-09-23 measurement, not a new run.
- **duckpeakwhere measured commit:** `r measured`, signed DuckHTS 1.5.2, duckdb-wasm
  1.5.4 (`wasm_eh`, one thread), W2 files over same-origin HTTP.
- **Inputs:** full GENCODE vM25 basic and full ENCODE thymus narrowPeak files
  ENCFF714WDP, ENCFF979ULB, ENCFF478UYW, ENCFF853BYO, ENCFF674JZY. Compressed hashes
  were verified against `bench/manifest.json`. There are 221,308 peaks.
- **Settings:** promoter −1000/+1000, centres, all transcripts. The first redraw
  switches to bp; the second changes promoter upstream to 2000, retaining bp mode.
- **Driver:** `node bench/head-to-head-ours.mjs 6 benchmarks/head-to-head-ours.json`.
  It submits the real app form, waits for rendered results, and uses the same session
  on redraws. Startup is navigation to the ready app and is outside first-analysis time.

## Seconds, median [min–max] of five runs

```{r timing}
rows <- data.frame(
  step = c("App startup", "First analysis (Run → results)",
           "Redraw: centres → base pairs", "Redraw: promoter upstream 2000 (bp)"),
  upstream = c("—", stat(col(up, function(r) r$run)),
               stat(col(up, function(r) r$redraw$mode_bp)), stat(col(up, function(r) r$redraw$promoter_2000))),
  duckpeakwhere = c(stat(col(ours, function(r) r$startup)), stat(col(ours, function(r) r$run)),
                    stat(col(ours, function(r) r$redraw$mode_bp)), stat(col(ours, function(r) r$redraw$promoter_2000))))
knitr::kable(rows, align = "lrr")
```

Upstream includes a 300 ms redraw debounce. duckpeakwhere measures submission through
rendering, with no debounce. Neither driver observed off-origin requests. The bp
redraw is under one second; the promoter redraw is **not**: it rebuilds categories and
the cgranges partition. The first analysis remains slower than upstream's 4.35 s.
The full feature read remains a [DuckHTS #249](https://github.com/RGenomicsETL/duckhts/issues/249)
work item; there is no application-side parser or projection workaround.

```{r before-after}
knitr::kable(data.frame(
  step = c("First analysis", "Mode redraw", "Promoter redraw"),
  before = c(stat(col(before, function(r) r$run)), stat(col(before, function(r) r$redraw$mode_bp)),
             stat(col(before, function(r) r$redraw$promoter_2000))),
  after = c(stat(col(ours, function(r) r$run)), stat(col(ours, function(r) r$redraw$mode_bp)),
            stat(col(ours, function(r) r$redraw$promoter_2000)))), align = "lrr")
```

The before receipt is preserved at `71c4674:benchmarks/head-to-head-ours.json` (the
report identifies its source as `5fa746b`). Its driver called uncached `annotate`
without rendering, and reset to centre mode for the promoter measurement. The after
driver includes rendering and keeps bp mode, matching upstream. The promoter rows
therefore compare workflows, not identical settings or identical timing boundaries.

## Exact agreement

```{r agreement}
csv <- up[[which(vapply(up, function(r) !is.null(r$csv), logical(1)))[1]]]$csv
lines <- strsplit(csv, "\n")[[1]]
reference <- read.csv(text = paste(lines[!startsWith(lines, "#")], collapse = "\n"))
reference <- reference[startsWith(reference$label, "thymus_"), ]
categories <- c("promoter", "utr5", "utr3", "exon", "intron", "intergenic")
marks <- vapply(strsplit(reference$label, "_"), `[`, "", 2L)
for (run in c(ours, before)) for (i in seq_along(marks)) {
  mine <- vapply(categories, function(k) as.numeric(run$counts[[marks[i]]][[k]]), numeric(1))
  stopifnot(identical(mine, setNames(as.numeric(unlist(reference[i, categories], use.names = FALSE)), categories)))
}
knitr::kable(data.frame(file = marks, identical_categories = 6L, of = 6L))
```

All 30 centre counts match both immutable pre-port receipts on **every** timed run
and warm-up, including native replay. This is independent upstream peakwhere agreement;
it is not a claim that the separate W2 R oracle finished. That oracle's timeout remains
recorded in [performance.md](performance.md). PeakPeek expectations are still its own
SPEC and tests at `a05b661` ([a05b6615b1c51c41781b6a1f2081868fbc2b711e](https://github.com/seandavi/peakpeek/tree/a05b6615b1c51c41781b6a1f2081868fbc2b711e)), not outputs from this port.
The expected file's values were re-checked against that commit: the fixture expectations
against `examples/fixture.js` and the five thymus rows against `SPEC.md` §2. (An earlier
revision of this report cited a PeakPeek commit `4f91069` that does not exist; it was a
recording error, corrected here.)

## SQL decisions and measured effects

- **Segment labelling and centre counting keep cgranges.** The supplied native-1t W2
  comparison found segment labelling at 0.84 s versus 1.64 s for islands + ASOF,
  1.51 s for islands + genome-coordinate IEJoin, and 1.68 s for IEJoin without islands.
  Centre probes took 0.079 s versus ASOF's 0.116 s. These operator experiments were
  not repeated. CTE range-join plans with poor estimates are not adopted.
- **Parent decoding:** only values containing `%` call `url_decode`; the supplied
  exact-equality measurement was 0.69 → 0.62 s native. Plain and encoded multi-parent
  links have a regression fixture.
- **File reads and caching:** one peak scan supplies validation, both views, contig
  reports and past-end warnings. The supplied profile attributed 0.32 of 0.70 s of
  count time to redundant reads. Annotation tables and the partition survive redraws;
  only dependent tables are rebuilt. File selection changes scan only added/replaced
  peak files. SQL-trace tests check the invalidation boundaries.
- **Bp reduction:** one materialized overlap list per matched peak feeds five genic
  sums; intergenic is total width minus those sums. Cache the annotation contig set
  rather than scanning all features per file. A bounded W2 browser probe measured
  bp 1.255 → 0.286 s and promoter-bp 2.744 → 1.793 s for these two changes together;
  all category counts at both bp settings exactly matched the pre-change SQL. This
  is a regression comparison, not an independent oracle. Independent hand fixtures
  check bp semantics. EXPLAIN JSON shows list reductions/`UNGROUPED_AGGREGATE`, a
  contig-membership `HASH_JOIN`, and no lateral `INOUT_FUNCTION`/UNNEST.
- **Peek histogram:** the bin grid and accepted rows are materialized temp tables.
  EXPLAIN on the 30-bin/17-row fixture selects `BLOCKWISE_NL_JOIN` for the inclusive
  last-bin predicate. No speedup is claimed for this materialization. There is no
  new interval range join in Where.

## Threading

`node bench/threading.mjs 3 benchmarks/threading.json` replays identical captured app
SQL over the same local HTTP URLs in signed native DuckDB `r threads$results[[1]]$version`.
One warm-up and three measured runs per setting; setting order alternates. These are
first-analysis SQL times, not app redraws or native local-file I/O.

```{r threads}
engines <- c("native-1t", "native-nt")
knitr::kable(do.call(rbind, lapply(engines, function(engine) {
  runs <- Filter(function(r) r$engine == engine, threads$results)
  data.frame(engine = engine, threads = runs[[1]]$threads,
    seconds = stat(col(runs, function(r) r$seconds$total)),
    rss_MiB = stat(col(runs, function(r) r$rss_kb / 1024)))
})), align = "lrrr")
```

The deployed browser probe reports crossOriginIsolated = **`r threads$wasm$crossOriginIsolated`**,
SharedArrayBuffer = **`r threads$wasm$sharedArrayBuffer`**, platform **`r threads$wasm$platform`**, and
`r threads$wasm$threads` thread. Signed extensions remain enforced and there are no
off-origin requests. **Browser threading is not usable in this deployment as served.**
GitHub Pages cannot set COOP/COEP response headers. A same-origin `coi-serviceworker`
could establish isolation, but worker registration/reload behavior and signed
`wasm_threads` execution would need a separate browser benchmark. No service worker
or threaded bundle is adopted; native thread scaling does not establish a browser
speedup. Threaded-browser throughput is unverified.
