---
output: github_document
---

<!-- Render with `make performance`; timing receipts come from `Rscript bench/run.R`. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = "")
source("../bench/report.R")
results <- jsonlite::read_json("results.json", simplifyVector = FALSE)
```

# Performance comparison

[Issue #1](https://github.com/sounkou-bioinfo/duckpeakwhere/issues/1) compares
[duckpeakwhere](../README.md)'s SQL in duckdb-wasm and native DuckDB with
ChIPseeker `annotatePeak`. Sean Davis's peakwhere is not timed: its
`src/pipeline.js` is a stub. The independent base-R oracle is a correctness
reference, not a timing competitor. These are local measurements, not CI benchmarks.

## Headline

Seconds, **median [min–max] of five runs**, after one discarded warm-up. Total
means the two analysis phases, excluding engine startup; native totals sum DuckDB's
per-statement wall timers. See process wall times below for startup-inclusive native
and R measurements. ChIPseeker performs additional nearest-transcript annotation and
has different category rules, so this is not an equal-output speedup claim.

```{r headline}
knitr::kable(headline(results), row.names = FALSE)
```

W1 contains all 7,220 bundled mouse thymus chr19 peaks in five files, against
GENCODE M25 basic chr19 (mm10/GRCm38). W2 requests the full versions of those same
GENCODE and ENCODE files. A failed workload has no timing entries.

## Reproduce

From the repository root, with the app's vendor files staged, the DuckDB CLI,
`/usr/bin/time`, headless Chromium, and R packages ChIPseeker, txdbmaker,
GenomicFeatures, jsonlite, processx, BiocManager, knitr and rmarkdown installed:

```bash
node bench/stage.mjs W1
Rscript bench/run.R W1
Rscript bench/run.R W2
make performance
make readme
```

[`bench/manifest.json`](../bench/manifest.json) pins the signed native DuckHTS
archive and all six full-genome input URLs and SHA-256s. `bench/run.R` calls the
staging script before the workload; `node bench/stage.mjs W2` also stages W2 directly.
Staging verifies the compressed bytes on every invocation
and extracts to `duckhts.duckdb_extension` (the basename determines the init symbol).
Downloads stay in ignored `bench/.cache/`. [`results.json`](results.json) contains
per-run receipts, including the discarded warm-up, counts, source/input SHA-256s,
versions, native statement timings and `/usr/bin/time -v` output. Native reports
DuckHTS 1.5.2 in `duckdb_extensions()`. That version field is blank in wasm, whose
release identity is instead established by checking the staged artifact against
`duckhts-manifest.json`'s pinned SHA-256 before every run. Rendering reads these
receipts; it does not rerun engines or download data.

## Measurement contract

- One warm-up and five measured runs per workload and engine, in fixed order:
  wasm, native-1t, native-nt, ChIPseeker. Each native/R run is a fresh process;
  each wasm run is a fresh page (and browser). OS caches are not flushed. Engines
  run serially. Each engine has a 600-second wall limit, including warm-up.
- Browser timing uses `performance.now()` at phase boundaries in the actual
  `annotate()` call. Phase **partition** starts before annotation header reading
  and includes feature/category construction and the partition index. Phase
  **count** includes all five peak files, chromosome diagnostics, genome background,
  metadata queries and index destruction, exactly as in the app. Rendering and
  database/extension startup are excluded.
- The browser captures the SQL issued by `src/annotate.js`. Native replays that
  text, rather than maintaining a SQL copy. The three engines' statement/phase
  SHA-256s must match for each repetition. Native uses the **same local HTTP URLs**
  while the browser's server remains alive; thus path literals also match exactly.
  No remote requests occur during analysis. Native local-file performance is not
  measured. Whole-file DuckHTS reads use `scan_mode := 'sequential'`.
- Native **partition/count/total** sum CLI `.timer on` real seconds per statement
  (printed at millisecond precision; small statements can round to zero). CLI
  stdout is retained: it includes the deprecated-lambda warning from the app SQL.
  Only the non-empty priority/count result sets are parsed for the native gate.
  **process_wall** measures the entire `/usr/bin/time ... duckdb` invocation,
  including loading, transport, CLI JSON serialization, and shutdown. The native
  SQL thread setting is either 1 or the CLI default; it is recorded, not assumed.
- ChIPseeker **partition** is GFF3 → TxDb; **count** reads the five BED files,
  constructs one-base centres, calls `annotatePeak`, and tabulates categories.
  **total** sums those phases. Package loading is outside analysis total but inside
  **process_wall**, which covers the fresh Rscript invocation and receipt writing.
  Annotation feature caches are shared across the five files within one run only.
- Native and R peak RSS come from `/usr/bin/time -v`, over the whole process.
  Browser memory is **not measured**. The host is shared with unrelated processes,
  with no CPU affinity or exclusive reservation. This single-machine, fixed-order
  sample has no confidence intervals and does not isolate CPU kernels from I/O or startup.

## Correctness gates and measurements

No workload timing is published unless all of its SQL correctness gates pass.
Every run, including warm-up, is checked. W1 requires the six counts for each file
to equal `test/fixtures/thymus-expected.json` exactly, with zero unmatched peaks;
`test/oracle.R` must reproduce that fixture byte for byte. W2 requires exact count
and unmatched-peak agreement among all three SQL engines, across all repetitions,
and exact count agreement with the independent oracle if it finishes within
**120 seconds**. An oracle timeout is reported explicitly; an oracle error or a
count mismatch stops the workload. `test/oracle.R` accepts annotation, peak-directory
and output paths, using the same base-R algorithm as W1. A failed workload writes
a failure receipt and exits nonzero; `make performance` can then render the failure
without any partial timings. ChIPseeker is not required to match SQL: input totals,
recognized category labels, and repeatability are checked,
with dropped peaks reported separately. Its disagreement is never used to relax the
SQL gate.

The brute-force oracle has a chromosome-scope limit: `category_of()` assigns a
chromosome absent from its interval list to intergenic, while SQL excludes
annotation-unmatched peaks. W2 includes 23 such peaks. This is an additional reason
not to claim independent W2 validation: a completed oracle remains subject to exact
comparison, with no exception for those peaks. The oracle's chromosome rule is not
used as a substitute for peakwhere's tested exclusion rule.

```{r measurements, results='asis'}
for (w in names(results)) {
  r <- results[[w]]
  cat("\n### ", w, "\n\n", sep = "")
  cat("Measured/attempted: ", r$date, "\n\n", sep = "")
  if (!is.null(r$oracle)) {
    cat("Oracle: **", r$oracle$status, "**; limit ", r$oracle$limit_seconds, " seconds.\n\n", sep = "")
    if (r$oracle$status == "timeout") {
      cat("No independent counts were obtained.\n\n```text\n", r$oracle$command,
          "\nexit_status: ", r$oracle$exit_status,
          "\nstdout: ", if (nzchar(r$oracle$log)) r$oracle$log else "(empty)",
          "\nstderr: ", if (nzchar(r$oracle$stderr)) r$oracle$stderr else "(empty)", "\n```\n\n", sep = "")
    }
  }
  if (r$status != "passed") {
    cat("**No timings reported.** ", r$skipped, "\n\n```text\n", r$failure, "\n```\n", sep = "")
    next
  }
  cat(r$gate, "\n\nInput peaks: ", sum(unlist(r$runs[[1]]$engines$chipseeker$input_n)), ".\n\n", sep = "")
  chroms <- sort(unique(unlist(lapply(r$runs[[1]]$engines$wasm$result$results, `[[`, "unmatchedChroms"))))
  if (length(chroms) > 0L) cat("SQL unmatched contigs: `", paste(chroms, collapse = "`, `"), "`.\n\n", sep = "")
  cat(knitr::kable(timing_table(r), row.names = FALSE), sep = "\n")
  cat("\n\n")
  cat(knitr::kable(memory_table(r), row.names = FALSE), sep = "\n")
  cat("\n\n")
}
```

## ChIPseeker agreement

Configuration: ChIPseeker 1.48.0; TxDb constructed from the **same GFF3**;
`level = 'transcript'`; `tssRegion = c(-1000, 1000)`;
`genomicAnnotationPriority = c('Promoter', '5UTR', '3UTR', 'Exon', 'Intron', 'Intergenic')`;
`ignoreDownstream = TRUE`. A BED interval `[s,e)` becomes a one-base GRanges interval
at `s + (e-s) %/% 2 + 1` (one-based). There is no summit-based or whole-peak counting.

Relevant rule differences, checked in the installed ChIPseeker functions
`annotatePeak`, `getNearestFeatureIndicesAndDistances` and `getGenomicAnnotation`:

- Both use transcript-level features here; **gene-level TSS is not the chosen
  setting**. SQL assigns promoter if *any* eligible transcript's ±1 kb interval
  contains the centre. ChIPseeker assigns promoter from its selected nearest-feature
  distance. In 1.48.0, `ignoreDownstream=TRUE` selects the `follow()` candidate on
  unstranded genomic coordinates (with TSS-overlap handling), not the minimum
  absolute distance of the two flanking TSS candidates. This can change promoters.
- `ignoreDownstream` does **not** disable downstream output labels. ChIPseeker's
  category function still labels remaining intergenic centres within its default
  3 kb downstream distance. Those labels are explicitly folded into **intergenic**;
  the number folded is shown below. SQL has no downstream category.
- SQL uses explicit GFF3 five/three-prime UTR features. TxDb represents CDS/exon
  structure and ChIPseeker obtains UTRs from `fiveUTRsByTranscript` and
  `threeUTRsByTranscript`; the annotation representations need not coincide.
- SQL treats transcript spans as intronic after higher-priority intervals are
  removed. ChIPseeker uses exon-gap introns from TxDb. TxDb transcript recognition
  and validation can differ from SQL's rule (a transcript is a parent of an exon).
  Construction warnings and transcript counts are retained in the receipts.
- Exon and intron transcript IDs/numbers and promoter distance labels are collapsed
  to their six category names, not compared as strings. Neither engine filters
  protein-coding transcripts or requires peak strand. SQL normalizes contig names
  and reports unmatched chromosomes; ChIPseeker uses TxDb sequence names and may
  omit centres lacking usable features. Such omissions are reported, not imputed.

An untimed W2 diagnostic replay with `options(warn = 1)` reproduced the recorded
ChIPseeker counts and dropped totals exactly. Its 20 annotation warnings came from
`.merge_two_Seqinfo_objects`: the peak files contain random/unplaced contigs absent
from TxDb, while TxDb contains chrM (and, for some files, chrY) absent from the peaks.
The SQL-unmatched contigs are listed above. TxDb construction also warns that
stop-codon phase values are ignored and genome-version metadata is unavailable;
the GRCm38 identity here comes from the pinned source provenance, not TxDb metadata.

Differences below are **ChIPseeker − SQL**. These are marginal category counts,
not per-peak concordance: equal totals do not prove the same peaks received a category.
No individual discrepancy is claimed to have been causally decomposed.

```{r agreement, results='asis'}
for (w in names(results)) {
  r <- results[[w]]
  if (r$status != "passed") next
  cat("\n### ", w, "\n\n", sep = "")
  cat(knitr::kable(agreement_table(r), row.names = FALSE), sep = "\n")
  chip <- r$runs[[1]]$engines$chipseeker
  cat("\n\n")
  cat(knitr::kable(data.frame(Mark = names(chip$input_n),
    Input = unlist(chip$input_n), `SQL unmatched` = unlist(r$runs[[1]]$engines$wasm$unmatched),
    `ChIPseeker dropped` = unlist(chip$dropped),
    `Downstream folded to intergenic` = unlist(chip$downstream), check.names = FALSE), row.names = FALSE), sep = "\n")
  cat("\n\nSQL transcripts: ", r$runs[[1]]$engines$wasm$result$meta$transcripts,
      "; TxDb transcripts: ", chip$transcripts, ".\n\n", sep = "")
}
```

## Environment and provenance

Source hashes in the receipts identify the measured working-tree files; the Git
revision alone is the base commit, not a claim that uncommitted benchmark code was
already part of that commit. Input hashes identify the exact compressed files.

```{r environment, results='asis'}
for (w in names(results)) {
  r <- results[[w]]
  if (r$status != "passed") next
  engines <- r$runs[[1]]$engines
  cat("\n### ", w, "\n\nMeasured: ", r$date, "\n\n", sep = "")
  versions <- c(Node = trimws(r$environment$node),
    `duckdb-wasm package` = r$environment$duckdb_wasm,
    `wasm DuckDB` = engines$wasm$version, Chromium = engines$wasm$chromium,
    `wasm platform` = engines$wasm$platform,
    `native DuckDB` = engines$`native-1t`$version, DuckHTS = r$environment$duckhts$version,
    R = engines$chipseeker$R, Bioconductor = engines$chipseeker$bioconductor,
    unlist(engines$chipseeker$packages))
  cat(knitr::kable(data.frame(Component = names(versions), Version = unname(versions)), row.names = FALSE), sep = "\n")
  cat("\n\nThreads: wasm ", engines$wasm$threads, "; native-1t ", engines$`native-1t`$threads,
      "; native-nt ", engines$`native-nt`$threads, ". Signed DuckHTS loaded with ",
      "`allow_unsigned_extensions=false` in all SQL engines.\n\n", sep = "")
  cpu <- strsplit(r$environment$cpu, "\n")[[1]]
  cpu <- cpu[grepl("^(Model name|CPU\\(s\\)|Core\\(s\\) per socket|Socket\\(s\\)|Thread\\(s\\) per core):", cpu)]
  cat("```text\n", paste(c(cpu, unlist(r$environment$os), trimws(r$environment$kernel)), collapse = "\n"), "\n```\n\n", sep = "")
  cat(knitr::kable(data.frame(File = names(r$input_sha256), SHA256 = unlist(r$input_sha256)), row.names = FALSE), sep = "\n")
  cat("\n\n")
}
```
