---
title: "Queryable alignment evidence: real PBMC comparison"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,message=FALSE,warning=FALSE)
raw<-read.delim("evidence/real/raw.tsv")
groups<-split(raw,interaction(raw$engine,raw$operation,raw$records,raw$threads,drop=TRUE))
summary<-do.call(rbind,lapply(groups,function(d)data.frame(
  Engine=d$engine[[1]],Operation=d$operation[[1]],Records=d$records[[1]],Threads=d$threads[[1]],
  Status=paste(unique(d$status),collapse=" / "),
  `CLI ms: median [min, max]`=sprintf("%.1f [%.1f, %.1f]",median(d$elapsed_s)*1000,min(d$elapsed_s)*1000,max(d$elapsed_s)*1000),
  `Max RSS MiB`=round(max(d$peak_rss_bytes,na.rm=TRUE)/1024^2,1),
  `Prepared MiB`=if(all(is.na(d$prepared_bytes)))NA_real_ else round(median(d$prepared_bytes,na.rm=TRUE)/1024^2,1),
  `Output cells`=paste(unique(d$output_rows[!is.na(d$output_rows)]),collapse=" / "),
  `Count sum`=paste(unique(d$count_sum[!is.na(d$count_sum)]),collapse=" / "),check.names=FALSE)))
```

# Queryable alignment evidence: real PBMC comparison

DuckHTS reads standard BAM; DuckDB prepares inspectable alignment, CIGAR,
junction and evidence-family tables. Region and junction macros query that
serving cache without reparsing CIGAR. Raw and corrected tags remain distinct.

This product reproduces useful AIE capabilities with an explicit counting
contract, rather than inheriting every Gravlax policy. Annotation replay remains
the demo's separately scoped [compatibility surface](REPORT.md); it is not yet
an independently validated product surface.

## Count meanings

- **UMI labels:** exact selected UMI strings per cell and query. They are not
  physical-molecule estimates; a reused UMI in another locus remains one label.
- **Evidence families:** exact cell/UMI/contig/strand keys with transitively
  overlapping primary reference spans. No arbitrary 50 kb gap or implicit
  sequence-error absorption. These are inspectable evidence groups, not proof
  of one biological molecule each.
- **Gravlax:** its corrected barcode assignments and UMI-class abstraction.

The default trusts supplied CB and retains raw UR for annotation-ready evidence.
An explicit UB cache selects supplied corrected tags. Missing tags are audited,
not silently substituted. See the [counting contract](COUNTING_CONTRACT.md).

**Different count meanings prevent an equal-output speedup claim.** The graphs
compare costs of the declared products and preserve their returned counts.

## Actual PBMC input

Source: public 10x Cell Ranger 3.0.0 `pbmc_1k_v3`, a **4.79 GB BAM** with GRCh38
contig names. The complete source was acquired over HTTPS with its observed ETag
fixed, size checked, `samtools quickcheck` verified and SHA-256 recorded.

One source pass produces nested **1M / 2M / 4M genuine primary records**—no
read or barcode replication. These inputs retain mapped, nonsupplementary NH=1
records with all CB/UB/CR/UR tags present. They are restricted real-cohort
prefixes, not the entire cohort and not general multimapper performance.
Source and derived BAMs use an owned tmpfs cache; prepared states use the local
filesystem. The [protocol](REAL_BENCHMARK_PROTOCOL.md) declares these choices and
budgets before acquisition and measurement.

## Preparation

![Real PBMC preparation with one thread: the same BAM enters two different retained representations.](../../site/assets/charts/aie-real-prepare-t1-latency.svg)

![Real PBMC preparation with four threads.](../../site/assets/charts/aie-real-prepare-t4-latency.svg)

![One-thread maximum whole-process RSS for preparation.](../../site/assets/charts/aie-real-prepare-t1-rss.svg)

![Four-thread maximum whole-process RSS for preparation.](../../site/assets/charts/aie-real-prepare-t4-rss.svg)

Preparation includes CLI startup, BAM parsing, policy processing and persisting
state. The SQL cache retains raw observations and two explicit count surfaces;
Gravlax retains its dedicated molecular archive. Prepared bytes are reported
below, not treated as representation parity.

## Full-chromosome query

![One-thread full chr1 query latency. SQL labels, SQL families and Gravlax classes have different count meanings.](../../site/assets/charts/aie-real-region-full-t1-latency.svg)

![Four-thread full chr1 query latency; these are cost contrasts, not equal-output speedups.](../../site/assets/charts/aie-real-region-full-t4-latency.svg)

![One-thread whole-CLI peak RSS for the full chr1 query.](../../site/assets/charts/aie-real-region-full-t1-rss.svg)

![Four-thread whole-CLI peak RSS for the full chr1 query.](../../site/assets/charts/aie-real-region-full-t4-rss.svg)

The fixed-window query is `1:1-3000000`; the full query is `1:1-248956422`.
Three fresh processes per point include startup, state opening, evaluation and
writing every cell count. There are no five-second padding loops. Short query
latencies are CLI diagnostics, not a STYLE timing verdict. Whiskers are the
three-run range, not confidence intervals.

## Validation and biological differences

The product passes an independent scalar CIGAR cursor and pairwise interval-graph
oracle under both tag policies. Tests cover later introns, D versus N, strand and
contig separation, transitive overlaps, GEM-group identity, missing tags and
primary filtering and variable-length leading-A UMI labels. Real raw-label counts
on the 4M-record input's full-chr1 and 3 Mb queries also match independent
samtools/awk. Four semantic mutants—skip cursor, 50 kb merging, strand
loss and secondary inclusion—disagree with that oracle; restored SQL passes.

The [real 109,271-record pilot](evidence/real-pilot/README.md) has 653 cell-count
differences when using supplied CB/UB. Selecting CB/raw UR reduces disagreement
with Gravlax to **one cell** (28 versus 29). That cell has three raw barcode
variants in the input. This narrows the remaining investigation; it does not
justify relabelling the entire discrepancy as intentional or demanding blind
upstream emulation.

The raw real-comparison rows include output cell counts and count sums for every
query. Returned counts must remain visible alongside performance.

## Exact data and limitations

```{r measurements}
knitr::kable(summary,row.names=FALSE)
```

[Raw rows, logs, source snapshots, input/runtime hashes and acquisition receipts](evidence/real/)
are preserved. Each process has a 180-second wall ceiling and 4 GiB RSS ceiling;
DuckDB uses 2 GiB buffer and 2 GiB spill limits. Prepared state is capped at
4 GiB. Failures remain in the table; no retry replaces them.

This is real-input evidence, not full product completion. Whole-cohort annotation
performance, a general multimapper workload and full STYLE qualification remain
unverified. The purpose-built original can still win: compare its preparation,
archive size and restricted-window cost with the richer SQL evidence cache.

## Reproduce

```bash
cd demos/aie
./check_evidence.sh
./derive_real.sh /path/to/pbmc_1k_v3.bam /new/ladder
Rscript --vanilla measure_real.R /new/ladder work/new-real-comparison
# One reusable product cache, with explicit tag policy:
./prepare_evidence.sh /path/to/input.bam work/input.duckdb sample UR 4
duckdb work/input.duckdb -c "SELECT * FROM aie_region_families('1',1,3000000);"
```

`DUCKDB` and `DUCKHTS_EXTENSION` select recorded runtime artifacts. Use new output
directories; preserve previous receipts. Render this report from committed
evidence, not by rerunning engines. From the repository root,
`Rscript scripts/plot-benchmarks.R` regenerates the static SVGs.
